Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-18 paper.tex
Minor edits

% Paragraph 2, what do we plot, specifically?
In this section we investigate the patterns of divergence between
the parents of recombination nodes.
We focus on the Long ARG because  it covers time
periods where recombination is known to have had a significant
impact, and includes
[HOW MANY?]
samples designated to [HOW MANY] Pango X lineages (i.e., those inferred
to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
further analysis of Pango recombinant lineages in the Long and Wide ARGs).
Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
plotted against the divergence between these parents (i.e., the
total branch length from the parents to their MRCA in the trees to
the immediate left and right of the breakpoint).
As in Section~\ref{sec:breakpoint_intervals}, we exclude recombination nodes
that are ancestral to only one sample from this analysis.
Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all recombinants that
are ancestral to more than one sample, and
Figure~\ref{fig:recomb_mrcas}B plots a subset which pass some further
reliability filters. Specifically, only recombination nodes that
[FILL ME IN].
This reduces the number of recombination nodes
to 347 and breakpoints to 358.
Note that each plotted point
represents a breakpoint, so that a recombination node with more than one
breakpoint (e.g. with 3 or more parents, comprising ${\sim}10 \%$ of the
recombination nodes in Figure ~\ref{fig:recomb_mrcas}A and ${\sim}3 \%$ in
Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
It is important to note here that time estimates in these figures may be unreliable,
since nodes other than samples are dated somewhat crudely in the current version
of \texttt{sc2ts} (Section~\ref{sec:node_dating}).

@@

% Paragraph 4, what do we say about the divergence times?
The histograms in Figure~\ref{fig:recomb_mrcas} show that there is a
broad range in divergance times among recombinant parents.
We can also see that the majority of recombinants arise
from parents that diverged 10 to 30 weeks
previously. However, there is also a peak of recombinants that diverged
${\sim}95$ weeks prior to the event: these correspond to recombinants whose
parental common ancestor traces to early 2020, including both Delta-Delta and
Delta-Omicron recombinants
(Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
Figure~\ref{fig:recomb_mrcas_voc_breakdown} also shows that, of the recombinant
parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
are within-Delta and within-Omicron recombinants, and that Omicron and
Delta are the variants associated with the most recombination. This may
reflect either sampling intensity, the prevalence of cases (which increases the
chance of coinfection and recombination), or possible heterogeneity in
recombination probabilities among lineages.

@@

\caption{\label{fig:pango-simple-origin-graph} Non-nested
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. Genomic regions inherited by the recombination node are shown
corresponding to the maximum inferred breakpoint position from the HMM process.
(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

@@

The classification of originating recombination events is dependent on
accurate designation of Pango lineages to samples. It is therefore important
to note that if the GISAID Pango designations are used, many of the samples
marked here as XAB are reclassified as BA.2 (and XAG becomes fully monophyletic,
see supplementary figure \ref{fig:pango_XAG_gisaid_graph}). This is an
independent confirmation of the uncertainty in designation of these XAG-related
samples.

@@

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. 15 Pango X lineages are inferred to have multiple
recombinant origins, such that their samples trace to >1 most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e. there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend (but in XJ and XU there are too few Pango X samples to decide on a ``main'' clade:
see Supplementary Table~\ref{tab:pango-recombinants}).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
the 4 samples labelled XD by Nextclade in the long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated maximum position of 21641 basepairs
(and minimum of 21619) with the insertion end occurring at a maximum position of 25584
(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent Deltacron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see Supplementary
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).

A similar pattern is seen for several of the origins inferred for the Pango
XN lineage (originally defined on the basis of sequences from the UK). After removing
singleton recombinants and the descendants of the erroneously recombinant
Denmark/DCGC-281594/2021 strain (which, however, includes 53 XN-designated samples)
the Long ARG contains four independent origins for Nextclade-designated XN
samples (Supplementary Table~\ref{tab:multiple_origins_table}). The oldest
of these four origins has 3 mutations above the sample that caused the
recombination to be inferred. It therefore seems likely that we have not managed
to sample an XN strain that is very close to the true recombinant. This may
lead to `texttt{sc2ts} mistakenly inferring multiple recombination events when there

@@

a recombinant outbreak
has taken a "true" XE sample and mistakenly inferred an additional, spurious origin. [WHY? - this
is presumably because it has changed too much from the nearest in the main XE cluster - need to discuss the
weighting of 3 mutations for one recombination here. NB: the number
of nucleotide differences between the closest XE in the two clusters is 11: see the
Pango-recombinant-analysis workbook]

\subsubsection{Nested recombinant origins}

@@

As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves multiple non-independent (``nested'') recombination events
(marked with asterisks in Supplementary Table~\ref{tab:pango-recombinants}; the most
extensive is for XAB). The earliest example, present in both the Wide and Long ARGs, is XB
The Logn ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted in blue). A similar
pattern is seen when examining XB in the Wide ARG (but as discussed previously,
in the Wide ARG, XB exhibits multiple nested recombinations, some with many XB
descendants).

>>>>

% Paragraph 2, what do we plot, specifically?
In this section we investigate the patterns of divergence between
the parents of recombination nodes.
We focus on the Long ARG because  it covers time
periods where recombination is known to have had a significant
impact, and includes
[HOW MANY?]
samples designated to [HOW MANY] Pango X lineages (i.e., those inferred
to have recombinant ancestry; see Section~\ref{sec:pango_x_lineages} for
further analysis of Pango recombinant lineages in the Long and Wide ARGs).
Figure~\ref{fig:recomb_mrcas} shows the estimated date of the most recent
common ancestor (MRCA) of the parent nodes for each recombination breakpoint,
plotted against the divergence between these parents (i.e., the
total branch length from the parents to their MRCA in the trees to
the immediate left and right of the breakpoint).
As in Section~\ref{sec:breakpoint_intervals}, we exclude recombination nodes
that are ancestral to only one sample from this analysis.
Figure~\ref{fig:recomb_mrcas}A plots breakpoints for all recombinants that
are ancestral to more than one sample, and
Figure~\ref{fig:recomb_mrcas}B plots a subset which pass some further
reliability filters. Specifically, only recombination nodes that
[FILL ME IN].
This reduces the number of recombination nodes
to 347 and breakpoints to 358.
Note that each plotted point
represents a breakpoint, so that a recombination node with more than one
breakpoint (e.g. with 3 or more parents, comprising ${\sim}10 \%$ of the
recombination nodes in Figure ~\ref{fig:recomb_mrcas}A and ${\sim}3 \%$ in
Figure ~\ref{fig:recomb_mrcas}B) will be represented by several points.
It is important to note here that some of the time estimates in these figures
come with a substantial degree of uncertainly,
since nodes other than samples are dated somewhat crudely in the current version
of \texttt{sc2ts} (Section~\ref{sec:node_dating}).

@@

% Paragraph 4, what do we say about the divergence times?
The histograms in Figure~\ref{fig:recomb_mrcas} show that there is a
broad range in divergence times among recombinant parents.
We can also see that the majority of recombinants arise
from parents that diverged 10 to 30 weeks
previously. However, there is also a peak of recombinants that diverged
${\sim}95$ weeks prior to the event: these correspond to recombinants whose
parental common ancestor traces to early 2020, including both Delta-Delta and
Delta-Omicron recombinants
(Figure~\ref{fig:recomb_mrcas_voc_breakdown}).
Figure~\ref{fig:recomb_mrcas_voc_breakdown} also shows that, of the recombinant
parents that fall into the Alpha, Delta, and Omicron VoC classes, the majority
are within-Delta and within-Omicron recombinants, and that Omicron and
Delta are the variants associated with the most recombination. This may
reflect either sampling intensity, the prevalence of cases (which increases the
chance of coinfection and recombination), or possible heterogeneity in
recombination probabilities among lineages.

@@

\caption{\label{fig:pango-simple-origin-graph} Non-nested
Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all five samples designated
as XA by Nextclade, together with their ancestral lineages, are shown outwards to the nearest
sampled viral genome; dotted lines show ARG continuation. Nodes are coloured by Nextclade
Pango designation; smaller symbols are non-sample nodes inserted by sc2ts, with
imputed Pango status. Genomic regions inherited by the recombination node are shown
corresponding to the maximum inferred breakpoint position from the HMM process.
(B) Equivalent plot for the 17 XAG samples in the long ARG, with abbreviated labelling.
(C) Equivalent plot for both origination events involving the four XD samples in the long ARG. Details of the mutations and sample node identities for all three plots are provided in supplementary Figures \ref{fig:pango_XA_gisaid_graph}, \ref{fig:pango_XAG_gisaid_graph}, and \ref{fig:pango_XD_gisaid_graph}.
} \end{figure}

@@

The classification of originating recombination events is dependent on
accurate designation of Pango lineages to samples. It is therefore important
to note that if the GISAID Pango designations are used, many of the samples
marked here as XAB are reclassified as BA.2 (and XAG becomes fully monophyletic,
see supplementary Figure \ref{fig:pango_XAG_gisaid_graph}). This is an
independent confirmation of the uncertainty in designation of these XAG-related
samples.

@@

\subsubsection{Independent multiple origins}
The \texttt{sc2ts} inference process has no pre-defined knowledge
of Pango X lineage assignments, and there is therefore no particular
requirement that all the samples assigned to a given lineage must trace back to
a single recombination event. 15 Pango X lineages are inferred to have multiple
recombinant origins, such that their samples trace to $>$1 most recent
recombination node in the ARG. Of these, 11 are cases where the
recombinants are independent rather than nested (i.e. there is no overlap in the list
of descendant samples for each recombination node). Most have a single ``main''
recombination event from which the majority of the corresponding recombinant samples
descend (but in XJ and XU there are too few Pango X samples to decide on a ``main'' clade:
see Supplementary Table~\ref{tab:pango-recombinants}).

Figure~\ref{fig:pango-simple-origin-graph}C shows a simple multiple-origin example, consisting of
the 4 samples labelled XD by Nextclade in the Long ARG. The left hand subgraph
(the origin of three XD samples, all sampled in France) has an earliest
sample with EPI\_ISL: 11222324 (dated 2022-02-26), while the right hand
subgraph has a single sample with EPI\_ISL~8514045 (sampled in Turkey and dated 2021-12-30).
Both involve an Omicron lineage being inserted into the middle of a Delta genome, but
the breakpoints in each case are slightly different: the start of the Omicron insertion
in the French samples has an estimated maximum position of 21641 basepairs
(and minimum of 21619) with the insertion end occurring at a maximum position of 25584
(and a minimum of 25470). In contrast, the Omicron insertion in the Turkish sample
is inferred to have occurred between positions 22578 (min 21619) and 24130 (min 23605).
The breakpoint difference, the different geographical locations, the time between the
samples, and the fact that the two samples differ at 23 nucleotide positions,
suggests that these may indeed represent true independent Deltacron recombinants.
The canonical XD definition is based entirely on samples from northern Europe, particularly
France (\url{https://github.com/cov-lineages/pango-designation/issues/444}) so it seems
plausible that the earlier Turkish sample has been mislabelled XD by Nextclade.
Indeed, GISAID does not label any of these samples XD (see Supplementary
Figure~\ref{fig:pango_XD_gisaid_graph} which gives exact mutations and sample identifiers).

A similar pattern is seen for several of the origins inferred for the Pango
XN lineage (originally defined on the basis of sequences from the UK). After removing
singleton recombinants and the descendants of the erroneously recombinant
Denmark/DCGC-281594/2021 strain (which, however, includes 53 XN-designated samples)
the Long ARG contains four independent origins for Nextclade-designated XN
samples (Supplementary Table~\ref{tab:multiple_origins_table}). The oldest
of these four origins has 3 mutations above the sample that caused the
recombination to be inferred. It therefore seems likely that we have not managed
to sample an XN strain that is very close to the true recombinant. This may
lead to \texttt{sc2ts} mistakenly inferring multiple recombination events when there

@@

a recombinant outbreak
has taken a ``true'' XE sample and mistakenly inferred an additional, spurious origin. [WHY? - this
is presumably because it has changed too much from the nearest in the main XE cluster - need to discuss the
weighting of 3 mutations for one recombination here. NB: the number
of nucleotide differences between the closest XE in the two clusters is 11: see the
Pango-recombinant-analysis workbook]

\subsubsection{Nested recombinant origins}
As well as cases where Pango X lineage origins are attributed to independent
recombination events, four Pango X lineages in the Long ARG have Nextclade-designated
samples whose ancestry involves multiple non-independent (``nested'') recombination events
(marked with asterisks in Supplementary Table~\ref{tab:pango-recombinants}; the most
extensive is for XAB). The earliest example, present in both the Wide and Long ARGs, is XB
The Logn ARG subgraph is plotted in Figure~\ref{fig:complex_origins_graph}, showing a recombination
between a B.1 sample and B.1.627 sample that leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are often not
identified as recombinants by Nextclade (plotted in blue). A similar
pattern is seen when examining XB in the Wide ARG (but as discussed previously,
in the Wide ARG, XB exhibits multiple nested recombinations, some with many XB
descendants).

@@
